Types from data: \textnormal{Making structured data first-class citizens in F\#}

### 2016-02-09 pldi-2016/paper.tex
Finish minor edits, draft stability

\begin{figure*}[t]
\noindent
\begin{equation*}
\begin{array}{rcll}
\ident{csh}(\sigma, \sigma) &=& \sigma & (\emph{eq})\\
\ident{csh}([\sigma_1], [\sigma_2]) &=& [\ident{csh}(\sigma_1, \sigma_2)] & (\emph{list}) \\
\ident{csh}(\bot, \sigma) = \ident{csh}(\sigma, \bot) &=& \sigma & (\emph{bot})\\
\ident{csh}(\kvd{null}, \sigma) = \ident{csh}(\sigma, \kvd{null}) &=& \addopt{\sigma}& (\emph{null}) \\
\ident{csh}(\kvd{any}, \sigma) = \ident{csh}(\sigma, \kvd{any}) &=& \kvd{any} & (\emph{top})\\
\ident{csh}(\kvd{float}, \kvd{int}) = \ident{csh}(\kvd{int}, \kvd{float}) &=& \kvd{float} & (\emph{num})\\
\ident{csh}(\sigma_2, \kvd{nullable}\langl \hat{\sigma_1} \rangl) = \ident{csh}(\kvd{nullable}\langl \hat{\sigma_1} \rangl, \sigma_2) &=& \addopt{\ident{csh}(\hat{\sigma_1}, \sigma_2)} & (\emph{opt})\\
\ident{csh}(\nu \; \{ \nu_1 \!:\! \sigma_1,  \ldots \;, \nu_n \!:\! \sigma_n \}, 
  \nu \; \{ \nu_1 \!:\! \sigma'_1, \; \ldots \;, \nu_n \!:\! \sigma'_n \}) &=& 
  \nu \; \{ \nu_1 \!:\! \ident{csh}(\sigma_1, \sigma'_1), \; \ldots \; , \nu_n \!:\! \ident{csh}(\sigma_n, \sigma'_n) \} & (\emph{recd}) \\
\ident{csh}(\sigma_1, \sigma_2) &=& \kvd{any} & (\emph{any})\\
\end{array}
\end{equation*}
%
\begin{equation*}
\quad\qquad
\begin{array}{rcll}
 \addopt{\hat{\sigma}} &\narrow{=}& \kvd{nullable}\langl\hat{\sigma}\rangl &(\textnormal{non-nullable shapes})\\
 \addopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\qquad\qquad
\begin{array}{rcll}
 \dropopt{\kvd{nullable}\langl\hat{\sigma}\rangl} &\narrow{=}& \hat{\sigma} &(\textnormal{nullable shape})\\
 \dropopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\end{equation*}

\caption{The common preferred shape relation}
\label{fig:subtyping-cst}
\end{figure*}

@@

Given two ground shapes, the \emph{common preferred shape} is the least upper bound of the
shape with respect to the preferred shape relation. The least upper bound prefers records, 
which is important for usability as discussed earlier (\S\ref{sec:providers-xml}).

\begin{definition}
A \emph{common preferred shape} of two ground shapes $\sigma_1$ and $\sigma_2$ is a shape 
$\ident{csh}(\sigma_1, \sigma_2)$ obtained according to Figure~\ref{fig:subtyping-cst}. 
The rules are matched from top to bottom.
\end{definition}

\noindent
The fact that the rules of \ident{csh} are matched from top to bottom resolves the ambiguity 
between certain rules. Most importantly (\emph{any}) is used only as the last resort.
When finding a common shape of two records (\emph{recd}) we find common preferred shapes of their 
respective fields. The field labels of two record types can always be made to match, through the 
shape inference from sample (\S\ref{sec:formal-inferval}), \emph{before} finding a common preferred 
shape using $\ident{csh}$.

@@

The \ident{csh} function finds the unique least upper bound of the partially ordered 
set of ground shapes (Lemma~\ref{thm:lub}).

\begin{lemma}[Least upper bound]
\label{thm:lub}
For ground $\sigma_1$ and $\sigma_2$, if $\sigma_1 \triangledown \sigma_2 \vdash \sigma$ then $\sigma$ is a least upper bound by $\sqsupseteq$.
\end{lemma}

\begin{figure}[!h]
\vspace{0em}  
\noindent
\begin{equation*}
\begin{array}{rclcrclcrcl}
 \semalt{i} &\narrow{=}& \ident{int} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} && \semalt{\kvd{true}} &\narrow{=}& \ident{bool}\\
 \semalt{f} &\narrow{=}& \ident{float} && \semalt{s} &\narrow{=}& \ident{string} && \semalt{\kvd{false}}  &\narrow{=}& \ident{bool}\\
\end{array}
\end{equation*}
\noindent
\vspace{-0.5em}
\begin{equation*}
\begin{array}{l}
 \semalt{[d_1; \ldots; d_n]} = [\semalt{d_1, \ldots, d_n}]
 \\[0.5em]
 \semalt{\nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}_{\rho_i}} =\\[0.1em]
 \qquad\nu~\{ \nu_1:\semalt{d_1}, \ldots, \nu_n :\semalt{d_n}, \addopt{\theta(\rho_i)} \}
 \\[0.5em]
 \semalt{d_1, \ldots, d_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
 \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{d_i} \vdash \sigma_i \\[0.6em]
 \textnormal{Choose minimal}\ \theta\ \textnormal{by ordering}\ \sqsubseteq\ \textnormal{lifted over substitutions} \\[0.1em]
\end{array}
\end{equation*}
\caption{Shape inference from sample data}
\label{fig:shape-inference}
\vspace{-0.5em}
\end{figure}

\begin{proof}
By induction over the definition of $\ident{csh}$. Note that the algorithm only infers the top shape 
\kvd{any} when on of the shapes is the top shape (\emph{top}) or when there is no other option 
(\emph{any}); a nullable shape is introduced in $\addopt{-}$ only when no non-nullable shape can
be used (\emph{null}), (\emph{opt}).
\end{proof}

@@

We now specify how we obtain the shape from data. As clrified later (\S\ref{sec:impl-parsing}), 
we represent JSON, XML and CSV documents using the same first-order \emph{data} value:
%
\begin{equation*}
\begin{array}{lcl}
 d &\narrow{=}& i \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
   &\narrow{|}& [d_1; \ldots; d_n] \lsep \nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}
\end{array}
\end{equation*}
%
The definition includes primitive values ($i$ for integers, $f$ for floats
and $s$ for strings) and \kvd{null}. A collection is written as a 
list of values in square brackets. A record starts with a name $\nu$, followed by a 
sequence of field assignments $\nu_i \mapsto d_i$.

Figure~\ref{fig:shape-inference} defines a mapping $\semalt{d_1,\ldots,d_n}$ which turns a 
collection of sample data $d_1, \ldots, d_n$ into a shape $\sigma$. Before applying $S$, we assume 
each record in each $d_i$ is marked with a fresh row inference variable $\rho_i$.

We then choose a ground, minimal, global substitution $\theta$ for row variables. The substitution
is minimal in the sense that it introduces as few new fields as possible, but no additional fields, 
and the newly introduced fields have the most specific shape. Because $\rho_i$ variables represent 
potentially missing fields, the $\addopt{-}$ operator from Figure~\ref{fig:subtyping-cst} is applied 
to all types in the vector.

@@

@@

\begin{figure*}
\noindent  
\begin{equation*}
\begin{array}{c}  
\inference
  {~}
  {\hspace{-0.5em}L; \Gamma \vdash d : \ident{Data}\hspace{-0.5em}}
~~
\inference
  {~}
  {\hspace{-0.5em}L; \Gamma \vdash i : \ident{int}\hspace{-0.5em}}
~~
\inference
  {~}
  {\hspace{-0.5em}L; \Gamma \vdash f : \ident{float}\hspace{-0.5em}}
~~
\inference
  {L; \Gamma, x : \tau_1 \vdash e : \tau_2}
  {L; \Gamma \vdash \lambda x.e : \tau_2}
~~
\inference
  {L; \Gamma \vdash e_2 : \tau_1 &
   L; \Gamma \vdash e_1 : \tau_1 \rightarrow \tau_2}
  {L; \Gamma \vdash e_1~e_2 : \tau_2}
\\[1.75em]
\inference
  {L; \Gamma \vdash e : \ident{Data}}
  {L; \Gamma \vdash \ident{hasShape}(\sigma, e) : \ident{bool}}
\quad
\inference
  { L; \Gamma \vdash e : \tau & \tau \in \{ \ident{int}, \ident{float} \} }
  { L; \Gamma \vdash \ident{convFloat}(\sigma, e) : \ident{float} }
\quad
\inference
  { L; \Gamma \vdash e_1 : \ident{Data} & L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
  { L; \Gamma \vdash \ident{convNull}(e_1, e_2) : \ident{option}\langle\tau\rangle }
\\[1.75em]
\inference
  { L; \Gamma \vdash e : \ident{Data} \\
    \ident{prim}\in\{ \ident{int}, \ident{string}, \ident{bool} \} }
  { L; \Gamma \vdash \ident{convPrim}(\ident{prim}, e) : \ident{prim} }
\quad
\inference
  { L; \Gamma \vdash e_1 : \ident{Data} \\ L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
  { L; \Gamma \vdash \ident{convElements}(e_1, e_2) : \ident{list}\langle\tau\rangle } 
\quad
\inference
  { L; \Gamma \vdash e_1 : \ident{Data} \\ L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
  { L; \Gamma \vdash \ident{convField}(\nu, \nu', e_1, e_2) : \tau }
\\[1.75em]
\inference
  {L; \Gamma \vdash e : C \\ \kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
  {L; \Gamma \vdash e.N_i:\tau_i}
\qquad
\inference
  {L; \Gamma \vdash e_i : \tau_i & \kvd{type}~C(x_1:\tau_1, \ldots, x_n:\tau_n) = \ldots \in L}
  {L; \Gamma \vdash \kvd{new}~C(e_1, \ldots, e_n):C}
\end{array}
\end{equation*}

@@

More formally, given a set of sample documents, code using the provided type is guaranteed to work if 
the inferred shape of the input is preferred with respect to the shape of any of the samples. Going back to 
\S\ref{sec:inference-subtyping}, this means that:
%
\begin{itemize}
\item[--] Input can contain smaller numerical values (e.g., if a sample contains float, the input can contain an integer)
\item[--] Records in the input can have additional fields
\item[--] Records in the input can have fewer fields than some of the records in the sample 
  document, provided that the sample also contains records that do not have the field 
\item[--] When a labelled top type is inferred from the sample, the actual input can also contain any other value,
  which implements the open world assumption  
\end{itemize}
%
The following lemma states that the provided code (generated in Figure~\ref{fig:tp-generation})
works correctly on an input $d'$ that is a subshape of $d$. More formally, the provided
provided expression (with input $d'$) can be reduced to a value and, if it is a class,
all its members can also be reduced to values.

\begin{lemma}[Correctness of provided types]
\label{thm:tp-correctness}
Given sample data $d$ and an input data value $d'$ such that $\semalt{d} \sqsupseteq \semalt{d'}$
and provided type, expression and classes $\tau, e, L = \sem{\semalt{d}}$, 
then $L, e~d' \reduce^{*} v$ and if $\tau$ is a class ($\tau=C$) then for all members $N_i$ of the 
class $C$, it holds that $L, (e~d').N_i \reduce^{*} v$.
\end{lemma}
\begin{proof}
By induction over the structure of $\sem{-}$. For primitives, the conversion functions accept all subtypes.
For other cases, analyze the provided code to see that it can work on all subtypes (for example~\ident{convElements}
works on \kvd{null} values, \ident{convFloat} accepts an integer); for labelled top types and
optional record fields, the \ident{hasShape} operation is used to guaranteed the correct shape at runtime.
\end{proof}

@@

For all inputs $d'$ such that $\sigma \sqsupseteq \semalt{d'}$ and all expressions $e'$
(representing the user code) such that $e'$ does not contain primitive operations $op$ as a sub-expression and
$\emptyset; y\!:\!\tau \vdash e'\!:\!\tau'$, it is the case that $L, e[y\leftarrow e'~d'] \reduce^{*} v$ for some value $v$ and
also $\emptyset; \vdash v : \tau'$.
\end{theorem}
\begin{proof}
We discuss the two parts of the proof separately as type preservation (Lemma~\ref{thm:rs-preservation})
and progress (Lemma~\ref{thm:rs-progress}).
\end{proof}

@@

\noindent
The progress lemma states that evaluation of a well-typed program does not reach an undefined state. 
This is not a problem for the Standard ML \cite{sml} subset and object-oriented subset \cite{fwjava} of the calculus. The 
problematic part are the dynamic data operations (Figure~\ref{fig:ff-reduction}). Given a data
value (of type \ident{Data}), the reduction can get stuck if the value does not have a structure 
required by a specific operation.

@@

The F\# Data library has been widely adopted by users and is one of the most downloaded
F\# libraries.\footnote{At the time of writing, the library has over 80,000 downloads on NuGet 
(package repository), 1,821 commits and 44 contributors on GitHub.} A practical demonstration of 
development using the library can be seen in an attached screencast and additional documentation can be
found at \url{http://fsharp.github.io/FSharp.Data}.

@@

The \emph{relative safety} property does not guarantee safety in the same way as traditional
closed-world type safety, but it reflects the reality that programming with external data is
increasingly important \cite{age-of-web}. Type providers increase the safety of this kind of programming,
as seen earlier (\S\ref{sec:introduction}).

One common issue is the handling of schema change. With type providers, the sample is captured 
at compile-time. If the schema changes (so that the input is no longer related to the  
shape of the sample), the program can fail at runtime and developers have to handle the exception. 
The same problem happens when using weakly-typed code with explicit failure cases.

F\# Data can help discover such errors earlier. Our first example (\S\ref{sec:introduction})
points the JSON type provider at a sample using a live URL. This has the advantage that a re-compilation 
fails when the sample changes, which is an indication that the program needs to be updated to reflect the
change. Repeated type-checking of programs by continuous integration systems can be used to help 
detect such changes.

In general, XML, CSV and JSON data sources without explicit schema will necessarily require techniques akin
to those we have shown. However, some data sources provide explicit schema with schema versioning support 
(with meta-data about how the schema changed). For those, a type 
provider could be written to adapt automatically, but we leave this for future work.

@@

\paragraph{Reading XML documents.}
Mapping XML documents to structural values is more interesting. For each node, we
create a record. Attributes become record fields and the body becomes a field with a special
name:
%
{\small{
\begin{verbatim}
  <root id="1">
    <item>Hello!</item>
  </root>    
\end{verbatim}
}}
%
\noindent
This XML becomes a record \ident{root} with fields \ident{id} and $\bullet$ for the body. 
The nested element contains only the $\bullet$ field with the inner text. As with CSV, we
infer shape of primitive values:
%
\begin{equation*}
\ident{root}~\{ \ident{id} \mapsto 1, \bullet \mapsto [ \ident{item}~\{ \bullet \mapsto \str{Hello!} \}] \}
\end{equation*}
%
When generating F\# types for XML documents, we also lifts the members nested under the $\bullet$ 
field into the parent type to simplify the structure of the generated type.

The XML type provider also includes an option to use \emph{global inference}. In that case, 
the inference from values (\S\ref{sec:formal-inferval}) unifies the shapes of \emph{all} records with the 
same name. This is useful because, for example, in XHTML all {\small\ttfamily <table>} elements
will be treated as values of the same type.

@@

When introducing type providers (\S\ref{sec:providers-sum}), we mentioned how F\# Data 
handles heterogeneous collections. This allows us to avoid inferring 
labelled top shapes in many common scenarios. In the earlier example, a sample collection 
contains a record (with \strf{pages} and \strf{page} fields) and a nested collection with values.

@@

\subsection{Capitalization} 
\label{sec:impl-naming}

@@

@@

>>>>

\begin{figure*}[t]
\noindent
\begin{equation*}
\begin{array}{rcll}
\ident{csh}(\sigma, \sigma) &=& \sigma & (\emph{eq})\\
\ident{csh}([\sigma_1], [\sigma_2]) &=& [\ident{csh}(\sigma_1, \sigma_2)] & (\emph{list}) \\
\ident{csh}(\bot, \sigma) = \ident{csh}(\sigma, \bot) &=& \sigma & (\emph{bot})\\
\ident{csh}(\kvd{null}, \sigma) = \ident{csh}(\sigma, \kvd{null}) &=& \addopt{\sigma}& (\emph{null}) \\
\ident{csh}(\kvd{any}, \sigma) = \ident{csh}(\sigma, \kvd{any}) &=& \kvd{any} & (\emph{top})\\
\ident{csh}(\kvd{float}, \kvd{int}) = \ident{csh}(\kvd{int}, \kvd{float}) &=& \kvd{float} & (\emph{num})\\
\ident{csh}(\sigma_2, \kvd{nullable}\langl \hat{\sigma_1} \rangl) = \ident{csh}(\kvd{nullable}\langl \hat{\sigma_1} \rangl, \sigma_2) &=& \addopt{\ident{csh}(\hat{\sigma_1}, \sigma_2)} & (\emph{opt})\\
\ident{csh}(\nu \; \{ \nu_1 \!:\! \sigma_1,  \ldots \;, \nu_n \!:\! \sigma_n \}, 
  \nu \; \{ \nu_1 \!:\! \sigma'_1, \; \ldots \;, \nu_n \!:\! \sigma'_n \}) &=& 
  \nu \; \{ \nu_1 \!:\! \ident{csh}(\sigma_1, \sigma'_1), \; \ldots \; , \nu_n \!:\! \ident{csh}(\sigma_n, \sigma'_n) \} & (\emph{recd}) \\
\ident{csh}(\sigma_1, \sigma_2) &=& \kvd{any} \quad(\textnormal{when}~\sigma_1\neq\nu\;\{ \ldots \}~\textnormal{or}~\sigma_2\neq\nu\;\{ \ldots \} ) & (\emph{any})\\
\end{array}
\end{equation*}
%
\begin{equation*}
\quad\qquad
\begin{array}{rcll}
 \addopt{\hat{\sigma}} &\narrow{=}& \kvd{nullable}\langl\hat{\sigma}\rangl &(\textnormal{non-nullable shapes})\\
 \addopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\qquad\qquad
\begin{array}{rcll}
 \dropopt{\kvd{nullable}\langl\hat{\sigma}\rangl} &\narrow{=}& \hat{\sigma} &(\textnormal{nullable shape})\\
 \dropopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\end{equation*}

\caption{The rules that define the common preferred shape function}
\label{fig:subtyping-cst}
\end{figure*}

@@

Given two ground shapes, the \emph{common preferred shape} is the least upper bound of the
shape with respect to the preferred shape relation. The least upper bound prefers records, 
which is important for usability as discussed earlier (\S\ref{sec:providers-xml}).
%
\begin{definition}
A \emph{common preferred shape} of two ground shapes $\sigma_1$ and $\sigma_2$ is a shape 
$\ident{csh}(\sigma_1, \sigma_2)$ obtained according to Figure~\ref{fig:subtyping-cst}. 
The rules are matched from top to bottom.
\end{definition}

\noindent
The fact that the rules of \ident{csh} are matched from top to bottom resolves the ambiguity 
between certain rules. Most importantly (\emph{any}) is used only as the last resort.
When finding a common shape of two records (\emph{recd}) we find common preferred shapes of their 
respective fields. The \ident{csh} function is not defined for two records with mis-matching 
fields, however, the fields can always be made to match, through the shape inference 
from sample (\S\ref{sec:formal-inferval}), \emph{before} finding a common preferred shape 
using $\ident{csh}$.

@@

When defined, \ident{csh} finds the unique least upper bound of the partially ordered 
set of ground shapes (Lemma~\ref{thm:lub}).

\begin{lemma}[Least upper bound]
\label{thm:lub}
For ground $\sigma_1$ and $\sigma_2$, if $\ident{csh}(\sigma_1, \sigma_2) \vdash \sigma$ then $\sigma$ is a least upper bound by $\sqsupseteq$.
\end{lemma}
\begin{proof}
By induction over the structure of the shapes $\sigma_1,\sigma_2$. Note that \ident{csh} only infers the top shape 
\kvd{any} when on of the shapes is the top shape (\emph{top}) or when there is no other option 
(\emph{any}); a nullable shape is introduced in $\addopt{-}$ only when no non-nullable shape can
be used (\emph{null}), (\emph{opt}).
\end{proof}

@@

We now specify how we obtain the shape from data. As clrified later (\S\ref{sec:impl-parsing}), 
we represent JSON, XML and CSV documents using the same first-order \emph{data} value:

\noindent
\begin{equation*}
\begin{array}{lcl}
  \\[-0.5em]
 d &\narrow{=}& i \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
   &\narrow{|}& [d_1; \ldots; d_n] \lsep \nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}
\end{array}
\end{equation*}
%
The definition includes primitive values ($i$ for integers, $f$ for floats
and $s$ for strings) and \kvd{null}. A collection is written as a 
list of values in square brackets. A record starts with a name $\nu$, followed by a 
sequence of field assignments $\nu_i \mapsto d_i$.

Figure~\ref{fig:shape-inference} defines a mapping $\semalt{d_1,\ldots,d_n}$ which turns a 
collection of sample data $d_1, \ldots, d_n$ into a shape $\sigma$. Before applying $S$, we assume 
each record in each $d_i$ is marked with a fresh row inference variable $\rho_i$. 
We then choose a ground, minimal substitution $\theta$ for row variables. Because $\rho_i$ variables 
represent potentially missing fields, the $\addopt{-}$ operator from Figure~\ref{fig:subtyping-cst} 
is applied to all types in the vector.

@@

\begin{figure}[!h]
\begin{equation*}
\begin{array}{rclcrclcrcl}
 \semalt{i} &\narrow{=}& \ident{int} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} && \semalt{\kvd{true}} &\narrow{=}& \ident{bool}\\
 \semalt{f} &\narrow{=}& \ident{float} && \semalt{s} &\narrow{=}& \ident{string} && \semalt{\kvd{false}}  &\narrow{=}& \ident{bool}\\
\end{array}
\end{equation*}
\noindent
\vspace{-0.5em}
\begin{equation*}
\begin{array}{l}
 \semalt{[d_1; \ldots; d_n]} = [\semalt{d_1, \ldots, d_n}]
 \\[0.5em]
 \semalt{\nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}_{\rho_i}} =\\[0.1em]
 \qquad\nu~\{ \nu_1:\semalt{d_1}, \ldots, \nu_n :\semalt{d_n}, \addopt{\theta(\rho_i)} \}
 \\[0.5em]
 \semalt{d_1, \ldots, d_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
 \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{d_i} \vdash \sigma_i \\[0.6em]
 \textnormal{Choose minimal}\ \theta\ \textnormal{by ordering}\ \sqsubseteq\ \textnormal{lifted over substitutions} \\[0.1em]
\end{array}
\end{equation*}
\caption{Shape inference from sample data}
\label{fig:shape-inference}
\vspace{-1em}
\end{figure}

@@

\begin{figure*}
\noindent  
\begin{equation*}
\begin{array}{c}  
\inference
  {~}
  {\hspace{-0.5em}L; \Gamma \vdash d : \ident{Data}\hspace{-0.5em}}
~~
\inference
  {~}
  {\hspace{-0.5em}L; \Gamma \vdash i : \ident{int}\hspace{-0.5em}}
~~
\inference
  {~}
  {\hspace{-0.5em}L; \Gamma \vdash f : \ident{float}\hspace{-0.5em}}
~~
\inference
  {L; \Gamma, x : \tau_1 \vdash e : \tau_2}
  {L; \Gamma \vdash \lambda x.e : \tau_2}
~~
\inference
  {L; \Gamma \vdash e_2 : \tau_1 &
   L; \Gamma \vdash e_1 : \tau_1 \rightarrow \tau_2}
  {L; \Gamma \vdash e_1~e_2 : \tau_2}
\\[1.75em]
\inference
  {L; \Gamma \vdash e : \ident{Data}}
  {L; \Gamma \vdash \ident{hasShape}(\sigma, e) : \ident{bool}}
\quad
\inference
  { L; \Gamma \vdash e : \ident{Data} & \tau \in \{ \ident{int}, \ident{float} \} }
  { L; \Gamma \vdash \ident{convFloat}(\sigma, e) : \ident{float} }
\quad
\inference
  { L; \Gamma \vdash e_1 : \ident{Data} & L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
  { L; \Gamma \vdash \ident{convNull}(e_1, e_2) : \ident{option}\langle\tau\rangle }
\\[1.75em]
\inference
  { L; \Gamma \vdash e : \ident{Data} \\
    \ident{prim}\in\{ \ident{int}, \ident{string}, \ident{bool} \} }
  { L; \Gamma \vdash \ident{convPrim}(\ident{prim}, e) : \ident{prim} }
\quad
\inference
  { L; \Gamma \vdash e_1 : \ident{Data} \\ L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
  { L; \Gamma \vdash \ident{convElements}(e_1, e_2) : \ident{list}\langle\tau\rangle } 
\quad
\inference
  { L; \Gamma \vdash e_1 : \ident{Data} \\ L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
  { L; \Gamma \vdash \ident{convField}(\nu, \nu', e_1, e_2) : \tau }
\\[1.75em]
\inference
  {L; \Gamma \vdash e : C \\ \kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
  {L; \Gamma \vdash e.N_i:\tau_i}
\qquad
\inference
  {L; \Gamma \vdash e_i : \tau_i & \kvd{type}~C(x_1:\tau_1, \ldots, x_n:\tau_n) = \ldots \in L}
  {L; \Gamma \vdash \kvd{new}~C(e_1, \ldots, e_n):C}
\end{array}
\end{equation*}

@@

More formally, given a set of sample documents, code using the provided type is guaranteed to work if 
the inferred shape of the input is preferred with respect to the shape of any of the samples. Going back to 
\S\ref{sec:inference-subtyping}, this means that:
%
\begin{itemize}
\item[--] Input can contain smaller numerical values (e.g., if a sample contains float, the input can contain an integer)
\item[--] Records in the input can have additional fields
\item[--] Records in the input can have fewer fields than some of the records in the sample 
  document, provided that the sample also contains records that do not have the field 
\item[--] When a labelled top type is inferred from the sample, the actual input can also contain any other value,
  which implements the open world assumption  
\end{itemize}
%
The following lemma states that the provided code (generated in Figure~\ref{fig:tp-generation})
works correctly on an input $d'$ that is a subshape of $d$. More formally, the provided
expression (with input $d'$) can be reduced to a value and, if it is a class,
all its members can also be reduced to values.

\begin{lemma}[Correctness of provided types]
\label{thm:tp-correctness}
Given sample data $d$ and an input data value $d'$ such that $\semalt{d'} \sqsubseteq \semalt{d}$
and provided type, expression and classes $\tau, e, L = \sem{\semalt{d}}$, 
then $L, e~d' \reduce^{*} v$ and if $\tau$ is a class ($\tau=C$) then for all members $N_i$ of the 
class $C$, it holds that $L, (e~d').N_i \reduce^{*} v$.
\end{lemma}
\begin{proof}
By induction over the structure of $\sem{-}$. For primitives, the conversion functions accept all subtypes.
For other cases, analyze the provided code to see that it can work on all subtypes (for example~\ident{convElements}
works on \kvd{null} values, \ident{convFloat} accepts an integer). Finally, for labelled top types,
the \ident{hasShape} operation is used to guaranteed the correct shape at runtime.
\end{proof}

@@

For all inputs $d'$ such that $\semalt{d'} \sqsubseteq \sigma$ and all expressions $e'$
(representing the user code) such that $e'$ does not contain any of the dynamic data operations $op$
and any \ident{Data} values as sub-expressions and $L; y\!:\!\tau \vdash e'\!:\!\tau'$, it is 
the case that $L, e[y\leftarrow e'~d'] \reduce^{*} v$ for some value $v$ and
also $\emptyset; \vdash v : \tau'$.
\end{theorem}
\begin{proof}
We discuss the two parts of the proof separately as type preservation (Lemma~\ref{thm:rs-preservation})
and progress (Lemma~\ref{thm:rs-progress}).
\end{proof}

@@

\noindent
The progress lemma states that evaluation of a well-typed program does not reach an undefined state. 
This is not a problem for the Standard ML \cite{sml} subset and object-oriented subset \cite{fwjava} of the calculus. The 
problematic part are the dynamic data operations (Figure~\ref{fig:ff-reduction}, Part I). Given a data
value (of type \ident{Data}), the reduction can get stuck if the value does not have a structure 
required by a specific operation.

@@

The F\# Data library has been widely adopted by users and is one of the most downloaded
F\# libraries.\footnote{At the time of writing, the library has over 100,000 downloads on NuGet 
(package repository), 1,844 commits and 44 contributors on GitHub.} A practical demonstration of 
development using the library can be seen in an attached screencast and additional documentation can be
found at \url{http://fsharp.github.io/FSharp.Data}.

@@

The \emph{relative safety} property does not guarantee safety in the same way as traditional
closed-world type safety, but it reflects the reality that programming with external data is
increasingly important \cite{age-of-web}. Type providers increase the safety of this kind of 
programming.

\paragraph{Representative samples.}
When choosing a representative sample document, the user does not need to provide a sample 
that represents all possible inputs. They merely need to provide a sample that is representative
with respect to data they intend to access. This makes the task of choosing a representative
sample easier.

\paragraph{Schema change.}
Type providers are invoked at compile-time. If the schema changes (so that inputs are no longer 
related to the shape of the sample used at compile-time), the program can fail at runtime and 
developers have to handle the exception. The same problem happens when using weakly-typed code 
with explicit failure cases.

F\# Data can help discover such errors earlier. Our first example (\S\ref{sec:introduction})
points the JSON type provider at a sample using a live URL. This has the advantage that a 
re-compilation fails when the sample changes, which is an indication that the program needs to be 
updated to reflect the change.

\paragraph{Richer data sources.}
In general, XML, CSV and JSON data sources without explicit schema will necessarily require 
techniques akin to those we have shown. However, some data sources provide explicit schema with 
schema versioning support. For those, a type provider could be written to adapt automatically, 
but we leave this for future work.

@@

\paragraph{Reading XML documents.}
Mapping XML documents to structural values is more interesting. For each node, we
create a record. Attributes become record fields and the body becomes a field with a special
name:
%
{\small{
\begin{verbatim}
  <root id="1">
    <item>Hello!</item>
  </root>    
\end{verbatim}
}}
%
\noindent
This XML becomes a record \ident{root} with fields \ident{id} and $\bullet$ for the body. 
The nested element contains only the $\bullet$ field with the inner text. As with CSV, we
infer shape of primitive values:
%
\begin{equation*}
\ident{root}~\{ \ident{id} \mapsto 1, \bullet \mapsto [ \ident{item}~\{ \bullet \mapsto \str{Hello!} \}] \}
\end{equation*}
%
The XML type provider also includes an option to use \emph{global inference}. In that case, 
the inference from values (\S\ref{sec:formal-inferval}) unifies the shapes of \emph{all} records with the 
same name. This is useful because, for example, in XHTML all {\small\ttfamily <table>} elements
will be treated as values of the same type.

\subsection{Providing idiomatic F\# types} 
\label{sec:impl-naming}

In order to provide types that are easy to use and follow the F\# coding guidelines,
we perform a number of transformations on the provided types that simplify their structure
and use more idiomatic naming of fields. For example, the type provided for the XML document in 
\S\ref{sec:impl-parsing} is:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Root}~=  \\
 \quad \kvd{member}~\ident{Id}~:~\ident{int} \\
 \quad \kvd{member}~\ident{Item}~:~\ident{string}
\end{array}
\end{equation*}
%
To obtain the type signature, we used the type provider as defined in Figure~\ref{fig:tp-generation}
and applied three additional transformations and simplifications:

\begin{itemize}
\item When a class $C$ contains a member $\bullet$, which is a class with further members, the
  nested members are lifted into the class $C$. For example, the above type \ident{Root}
  directly contains \ident{Item} rather than containing a member $\bullet$ returning a 
  class with a member \ident{Item}.

\item Remaining members named $\bullet$ in the provided classes (typically of primitive
  types) are renamed to \ident{Value}.

\item Class members are renamed to follow \ident{PascalCase} naming convention, when a 
  colision occurs, a number is appended to the end as in \ident{PascalCase2}. The provided
  implementation preforms the lookup using the original name.
\end{itemize}

Our current implementation also adds an additional member to each class that returns the
underlying JSON node (called \ident{JsonValue}) or XML element (called \ident{XElement}). 
Those return standard .NET or F\# representation of the value and can be used to dynamically 
access data not exposed by the type providers, such as textual values inside mixed-content XML elements.

@@

When introducing type providers (\S\ref{sec:providers-sum}), we mentioned how F\# Data 
handles heterogeneous collections. This allows us to avoid inferring 
labelled top shapes in many common scenarios. In the earlier example, a sample collection 
contains a record (with \strf{pages} field) and a nested collection with values.

@@

@@

One of the design motivations mentioned in Section~\ref{sec:providers} is that the inference
mechanism is predictable and stable. Informally this means that ...

@@

For the purpose of this section, assume conversion function \ident{int} and an error value
\ident{error} that propagates, i.e. $C[\ident{error}]\reduce\ident{error}$

\begin{theorem}[Stability of inference]

Given set of samples $d_1, \ldots, d_n$

And $\tau_1, e_1, L_1 = \sem{\semalt{d_1, \ldots, d_n}}$

And some code $L_1; x:\tau_1\vdash e : \tau$

~

Now, add $d_{n+1}$ and let
 $\tau_2, e_2, L_2 = \sem{\semalt{d_1, \ldots, d_n, d_{n+1}}}$

There is $e'$

such that $L_2; x:\tau_2\vdash e' : \tau$

and if for some $d$ it is the case that $e[x\leftarrow e_1~d] \reduce v$ then
also $e'[x\leftarrow e_2~d] \reduce v$

Obtained by changing

\begin{itemize}
\item[--]
$e$ to $\kvd{match}~e~\kvd{with}~\ident{Some}(v) \rightarrow v~|~\ident{None} \rightarrow \ident{wrong}$
\item[--]
$e$ to $e.\ident{Member}$ where $\ident{Member} = \tytagof(\sigma)$ for some $\sigma$
\item[--]
$e$ to $\ident{int}(e)$
\end{itemize}
\end{theorem}
\begin{proof}
